Sequential algorithms to split and merge ultra-high resolution 3D images

### 2017-08-01 paper.tex
Revised up until II.F

\begin{abstract}
  Splitting and merging data is a requirement for most parallel or
  distributed processing operations. Naive algorithms, however,
  perform very poorly to split or merge 3D blocks from ultra-high
  resolution images, due to important seek times. We propose and
  analyze two algorithms based on in-memory reconstruction of image
  parts. Cluster reads access image splits only once while Multiple
  reads minimize the accesses to the image. Evaluation on a
  3000x3000x3000 brain image shows that our algorithms perform
  similarly to the optimal configuration (slice merging) provided that
  enough memory is available, which solves our initial problem.  We
  conclude that splitting and merging large 3D images can be done
  efficiently based on simple data formats.
\end{abstract}

@@

In our context, split and merge relate to the same dual problem . We
focus here on merging for the sake of concision. Splitting algorithms
can be derived from merging by swapping reads and writes. Our goal
then is to merge a set of $n$ chunks into a single reconstructed 3D
image with $R$ voxels of size $b$. We assume that all chunks are of
identical size and that slices are squares and blocks are cubes.

Without loss of generality, we consider a file format where voxels are
written after alphanumerical sorting of their (z,y,x)
representation. That is, z slices are written one after the other, and
in every z slice, y rows are written one after the other. x is the
fastest-changing dimention.  NIfTI, the file format defined by the
Neuroimaging Informatics Technology Initiative
NIfTI\footnote{\url{https://nifti.nimh.nih.gov}} is such a format,
used by the vast majority of the neuroimaging community. In the
remainder, we use the term ``slice" for ``z slice".

@@

We adopt the following notations (see Figure\ref{fig:notations}):
\begin{itemize}
\item $R=D^3$: number of voxels in the reconstructed image.
\item $b$: number of bytes per voxel in the reconstructed image and chunks (in B).
\item $n$: number of chunks (blocks or slices).
\item $m$: amount of available memory (in B).
\item $m'$: amount of used memory (in B), $m'\leq m$.
\end{itemize}

For the sake of mathematical simplicity, we assume that the
reconstructed image and blocks are cubic and that they do not
overlap. Under this assumption, we also have the following relations:
\begin{itemize}
\item Number of slices in a block: $\sqrt[3]{\frac{R}{n}}=d$. This is
  also the number of rows and columns in a slice and the number of
  voxel in a row or column.
\item Number of blocks required to fill a row of the reconstructed
  image: $\sqrt[3]{n}$.
\end{itemize}

%\todo{Make a figure to illustrate the notations.}

@@

We use a simplistic disk model where a disk is characterized by its
read and write rates and its seek time. We do not distinguish between
access time and seek time. We also assume that seeks require a
constant amount of time, regardless of the position seeked to. In
practice, large variations would be expected, but modeling such
variations would inevitably lead to models specific to the hardware,
file system or operating system, which we intentionally avoid
here. Thus, our goal is to find algorithms that minimize the
\emph{number} of seek operations. Likewise, in modern systems, read
and write times are greatly impacted by caching occurring at several
levels, which we do not model here.

%% and its seek time,
%% a function $\sigma$ of the seek distance (in B). More precisely,
%% $\sigma$ would typically be in the order of $\beta$ when an opened
%% file is seeked to a random position, but it would be much smaller when
%% a file is seeked to a nearby position.
%% The disk model encompasses file systems, kernel modules and disk
%% controllers required to access the disk from a program run in user
%% space. In modern systems, this often includes caching at various
%% levels \todo{\cite{x}}, which we will model by zeroing $\alpha_r$ or
%% $\alpha_w$ values depending on the application.

@@

Algorithms~\ref{algo:naive-slices} and~\ref{algo:naive-blocks} are the
naive merging methods for slices and blocks.
\begin{algorithm}[h]
\caption{Naive merging from slices.}
\label{algo:naive-slices} 
\begin{algorithmic}
  \FOR{each slice}
    \STATE read slice
    \STATE write slice in reconstructed image
  \ENDFOR      
\end{algorithmic}
\end{algorithm}
\begin{algorithm}[h]
\caption{Naive merging from blocks.}
\label{algo:naive-blocks}
\begin{algorithmic}
  \FOR{each block}
    \STATE read block
    \STATE write block in reconstructed image
  \ENDFOR 
\end{algorithmic}
\end{algorithm}
Algorithms~\ref{algo:naive-slices} and~\ref{algo:naive-blocks}
actually have very different performance even though blocks and slices
have identical sizes. Since slices are stored contiguously in the
reconstructed image, the number of seeks in
Algorithm~\ref{algo:naive-slices} is only $2n$ as $n$ seeks are
required to read the slices and $n$ seeks are required to write them:
\begin{equation}
  N_\mathrm{slices} = 2n 
\end{equation}
However,
Algorithm~\ref{algo:naive-blocks} has to do extra seeks for each row
in each slice of each block. That is, the total number of seeks
performed by Algorithm~\ref{algo:naive-blocks} is:
\begin{equation*}
N_\mathrm{blocks} = n+nd^2 
\end{equation*}
or, using $R$ and $n$ as our main variables:
\begin{equation}
N_\mathrm{blocks} = n+n\left(\sqrt[3]{\frac{R}{n}}\right)^2
\end{equation}
In
practice, this difference could lead to a tremendous slowdown as we will show
later.

%%  For slices, the total merge time is modeled as follows:
%% \begin{equation}
%% T_\mathrm{slices} = \frac{bR}{\alpha_r}+\frac{bR}{\alpha_w}+2\beta n
%% \end{equation}
%% and for blocks:
%% \begin{equation}
%%   \begin{multlined}
%%     T_\mathrm{blocks} = \frac{bR}{\alpha_r}+\frac{bR}{\alpha_w}+ 2\beta n + \\
%%     n\left({\sigma(bR/2)} + d(d-1)\sigma\left((D-d)b\right)+d\sigma\left((D^2-d^2)b\right) \right)
%%     \end{multlined}
%% \end{equation}

\begin{figure}
  \centering
  \begin{minipage}[b]{0.3\columnwidth}
    \def\svgwidth{\columnwidth}
    \input{figures/svg/Notations.pdf_tex}
    \caption{Notations.}
    \label{fig:notations}
  \end{minipage}
  \quad \quad \quad
  \begin{minipage}[b]{0.3\columnwidth}
    \def\svgwidth{\columnwidth}
    \input{figures/svg/buffer.pdf_tex}
    \caption{Buffer used in cluster reads.}
    \label{fig:cluster-reads-buffer}
  \end{minipage}
\end{figure}

@@

Algorithm~\ref{algo:naive-slices} is a particular case of memory
buffering where the amount of available memory equals the maximum size
of a chunk. More buffering can be achieved when the amount of
available memory increases, as shown in
Algorithm~\ref{algo:buffered-slices}.
\begin{algorithm}[h]
  \caption{Buffered merging from slices}
  \label{algo:buffered-slices}
  \begin{algorithmic}[1]
    \STATE sorted\_slices = sort slices by increasing z values
    \STATE initialize buffer
    \FOR{i = 0 ; i \textless n ; i+=1 }
      \STATE slice = sorted\_slices[i]
      \IF{\texttt{sizeof}(buffer)+\texttt{sizeof}(slice) $\geq$ m}
        \STATE write buffer in reconstructed image
        \STATE clear buffer
      \ENDIF
      \STATE read slice and append it to buffer
    \ENDFOR
  \end{algorithmic}
\end{algorithm}
The number of seeks obtained with buffered slices is:
\begin{equation}
N_\mathrm{buff\_slices} =  n + \ceil*{\frac{bR}{m}}
\end{equation}
The next two algorithms attempt to generalize buffered slices to blocks.

@@

The main idea of cluster reads is to load multiple blocks in memory,
concatenate them in a buffer and write the buffer in the reconstructed
image. Seeking is reduced compared to naive block merging since
contiguous parts of the buffer will be written without seeking. In
cluster reads, a given block is accessed only once.

The buffer might be a slightly complex data structure, for instance an
associative array or a Python dictionary, capable of storing multiple
disjoint sequences of contiguous bytes without having to allocate
memory for the bytes between such
sequences. Figure~\ref{fig:cluster-reads-buffer} illustrates how the
buffer would fill up for the two first blocks in a reconstructed
image. It shows how the two first 4x4 blocks of a reconstructed image
would be arranged in the buffer.  White portions in the buffer are not
allocated.  The column-major order is used, similar to the NIfTI
format: i is the fastest changing dimension, followed by j, followed
by k.

@@

\begin{figure}
\centering
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case1-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case2-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case3.pdf_tex}
\caption{Memory-load configurations in cluster reads, leading to
  different number of seeks. Red blocks need seeking before each of
  their rows ($d^2$ seeks). Blue blocks need seeking before each of
  their slices ($d$ seeks in total). Green blocks need only a single
  seek. Grey, dashed, transparent blocks represent the contiguous
  memory loads and are added for the sake of visualization.}
\label{fig:cluster-reads}
\end{figure}
Our cluster reads algorithm focuses on the three memory load
configurations represented in Figure~\ref{fig:cluster-reads}, that is,
the amount of memory $m'$ used by the algorithm is rounded down to the
closest number of complete blocks (case 1), of complete block rows (case
2) or of complete block slices (case 3). This is in general reasonable
since adding an incomplete row to a set of complete ones multiplies
the number of required seeks by $d$, as illustrated in
Figure~\ref{fig:avoided-configurations}-Left. In some cases though,
rounding $m$ down to $m'$ might increase the number of required
memory load to a point that the overall number of seeks also
increases. Such cases are, however, slightly unusual and their
complete description requires extensive calculations involving modulo
arithmetic, which we felt were unwieldy to report here.
\begin{figure}
  \centering
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/incomplete-rows.pdf_tex}
\quad \quad \quad
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/overlap.pdf_tex}
\caption{Configurations that increase the number of seeks per
  memory-load and are thus deliberately avoided by cluster
  reads. Left: configuration with incomplete block rows (multiplies
  number of seeks by $d$). Right: configuration with block rows that
  overlap multiple block slices (multiplies number of seeks by $2$).}
\label{fig:avoided-configurations}
\end{figure}

Cluster reads are described in
Algorithm~\ref{algo:cluster-reads}. Function \texttt{switch} (line 3)
selects one of the three cases based on the amount of available memory
and the number of blocks. It returns \texttt{m'} and \texttt{case},
the identifier of the selected case. Function \texttt{check\_overlap}
(line 7) determines whether two blocks overlap multiple block slices
(case 2) or multiple block rows (case 1). For case 3 it always returns
true.  Function \texttt{sizeof} (line 8) returns the actual memory
used by its argument, including only its allocated segments in case
the argument is a buffer. 
\begin{algorithm}[h]
  \caption{Buffered merging of blocks with cluster reads}
  \label{algo:cluster-reads}
  \begin{algorithmic}[1]
    \STATE sorted\_splits = sort splits by increasing (z,y,x) values
    \STATE initialize buffer
    \STATE (m',case)=\texttt{switch}(m,n,R,b)
    \STATE old\_block = sorted\_splits[0]
    \FOR{i = 0 ; i\textless n ; i+=1}
      \STATE block = sorted\_splits[i]
      \STATE overlap = \texttt{check\_overlap}(block,old\_block,case)
      \IF{\texttt{sizeof}(buffer)+\texttt{sizeof}(block) $\geq$ m' \textbf{or} overlap=true}
      \STATE write buffer in reconstructed image
      \STATE clear buffer
      \STATE overlap = false
      \ENDIF
      \STATE read block and insert it in buffer
      \ENDFOR
  \end{algorithmic}
\end{algorithm}

The amount of memory used $m'$ is set as follows in each of the 3 cases:
\begin{equation*}
  m_1' = \frac{Rb}{n}\floor*{\frac{mn}{Rb}};
  m_2' = \frac{Rb}{\sqrt[3]{n}^2}\floor*{\frac{m\sqrt[3]{n}^2}{Rb}};
  m_3' = \frac{Rb}{\sqrt[3]{n}}\floor*{\frac{m\sqrt[3]{n}}{Rb}}
\end{equation*}
The number of seeks performed by cluster reads in each of the three cases is:
\begin{equation*}
  N^i_{\mathrm{CR}} = n + x_ib_i, \quad i \in \llbracket 1, 3\rrbracket,
\end{equation*}
where $x_i$ is the number of memory loads required to reconstruct the
image and $b_i$ is the number of seeks required to write a the memory
load. The first $n$ seeks in the equation are required to read all the blocks
once. According to Figure~\ref{fig:cluster-reads}, we have:
\begin{equation*}
  b_1=d^2=\sqrt[3]{\frac{R}{n}}^2 \quad ; \quad b_2=d=\sqrt[3]{\frac{R}{n}} \quad ; \quad b_3=1
\end{equation*}

Our algorithm avoids configurations where the memory load
overlaps multiple block slices in case 2 or multiple block rows in
case 1, as such overlaps multiply the number of required seeks
(see Figure~\ref{fig:avoided-configurations}-Right). Thus, the
numbers of memory loads required to reconstruct the image are:
\begin{equation*}
  x_1 = \ceil*{\frac{Rb}{\sqrt[3]{n}^2m_1'}}\sqrt[3]{n}^2;
  x_2 = \ceil*{\frac{Rb}{\sqrt[3]{n}m_2'}}\sqrt[3]{n};
  x_3 = \ceil*{\frac{Rb}{m_3'}}
\end{equation*}

@@

Multiple reads are shown in Algorithm~\ref{algo:multiple-reads}.  The
main idea of this algorithm is that blocks are read partially (line 9)
to ensure that the memory buffer only contains contiguous bytes, at
any time. Therefore, the buffer can be written continuously to the
reconstructed image, without seeking (line 13).  It should be noted
that a given block might be read multiple times, in different memory
loads.

@@

\begin{algorithm}[h]
  \caption{Buffered merging of blocks with multiple reads}
  \label{algo:multiple-reads}
  \begin{algorithmic}[1]
  \STATE sorted\_splits = sort splits by increasing (z,y,x) values
  \STATE start\_index = 0 ; end\_index=(m-1)
  \STATE write\_range = (start\_index, end\_index)
  \WHILE{end\_index \textless Rb}
    \STATE initialize buffer
    \FOR{split in sorted\_splits}
      \IF{split has voxels in write\_range}
        \STATE split\_data = read split
        \STATE in split\_data, extract the rows in write\_range
        \STATE insert rows in buffer
      \ENDIF
    \ENDFOR
    \STATE write buffer to reconstructed\_image
    \STATE start\_index = end\_index + 1 ; end\_index += m
  \ENDWHILE

@@

As illustrated in Figure~\ref{fig:multiple-reads-cases}, we assume that $m'$ represents an integer number $k$
of block rows (case 1, $k<\sqrt[3]{n}$), of complete rows (case 2,
$k<d$), of tile rows (case 3, $k<\sqrt[3]{n}$), of slices (case 4,
$k<d$) or of block slices (case 5, $k<\sqrt[3]{n}$). In each of these 5 cases, we define $v$ as follows:
\begin{equation*}
  v_1=db \text{ ; }  v_2=Db \text{ ; } v_3=Ddb \text{ ; } v_4=D^2b \text{ ; } v_5=D^2db
\end{equation*}
so that we have:
\begin{equation*}
k_i=\floor*{\frac{m}{v_i}} \quad \mathrm{and} \quad m_i'=k_iv_i, \quad i \in \llbracket 1,5 \rrbracket
\end{equation*}

The total number of seeks performed by multiple reads in case $i$ is:
\begin{eqnarray*}
  N^i_{\mathrm{MR}} &=& x_i + (x_i-1)b_i +b_i', \quad i \in \llbracket 1,5 \rrbracket\\
  &=& x_i \left(1+b_i\right)-b_i+b_i'
\end{eqnarray*}
where $x_i$ is the total number of required memory loads, $b_i$ is the
number of blocks accessed by the first ($x_i$-1) memory loads and
$b_i'$ is the number of blocks access by the last memory load. The
first $x_i$ seeks in the equation are required to write each memory
load (1 seek per memory load). We have:
\begin{equation*}
  x_i = \ceil*{\frac{Rb}{m_i'}}, \quad i \in \llbracket 1,5 \rrbracket
\end{equation*}
and:
\begin{equation*}
b_1=k_1 ; \quad b_2=\sqrt[3]{n} ; \quad b_3 = k_3\sqrt[3]{n} ; \quad b_4=\sqrt[3]{n}^2 ; \quad b_5=k_5\sqrt[3]{n}^2
\end{equation*}
and:
\begin{eqnarray*}
b_1'=\sqrt[3]{n}D^2\text{ mod }k_1; &b_2'=b_2;   & b_3' = \sqrt[3]{n}\left( \sqrt[3]{n}D\text{ mod }k_3\right)\\
                                   &b_4'=b_4;  & b_5'=\sqrt[3]{n}^2\left( \sqrt[3]{n}\text{ mod }k_5\right) 
\end{eqnarray*}
It gives the following expression for $N_\mathrm{MR}$:
\begin{equation*}
  N_\mathrm{MR} =
\begin{cases}
  \ceil*{\frac{Rb}{m_1'}}(k_1+1)-k_1 \\
  + \left( \sqrt[3]{n}D^2 \text{ mod } k_1 \right)& \text{\hspace*{-0.9em}if } d \leq \frac{m}{b} < D\\[0.4em]

@@

\begin{itemize}
\item In MR, no seeking withing the splits.
\item Avoid in-memory copies.
\end{itemize}

@@

For faster reading all the splits, before the merging process start,
we read all the split’s headers to get their position information and
calculate each row’s index in the reconstructed image. We only read the
split when there is a row in that writing range (available memory),
otherwise, we skip it.

@@

% Parallel algorithms
In a real system, high-resolution images are likely to be processed on
computing clusters, for instance using software from the Hadoop
project, in particular the Hadoop Distributed File
System~\cite{shvachko2010hadoop}. In this context parallel split-and-merge algorithms
will be required, since the various splits of a large image would be
uploaded to different disks concurrently. In the same vein,
``re-spliting'' algorithms would have to be devised in case a split
image need to be split in a different geometry. Cluster reads and
multiple reads could be used as starting points for such algorithms.

@@

>>>>

\begin{abstract}
  Splitting and merging data is a requirement for most parallel or
  distributed processing operations. Naive algorithms, however,
  perform very poorly to split or merge 3D blocks from ultra-high
  resolution images, due to important seek times. We propose and
  analyze two algorithms based on in-memory reconstruction of image
  parts. Cluster reads access image chunks only once while Multiple
  reads minimize the accesses to the image. Evaluation on a
  3000x3000x3000 brain image shows that our algorithms perform
  similarly to the optimal configuration (slice merging) provided that
  enough memory is available, which solves our initial problem.  We
  conclude that splitting and merging large 3D images can be done
  efficiently based on simple data formats.
\end{abstract}

@@

Split and merge relate to the same dual problem in our context. We
focus here on merging for the sake of concision. Splitting algorithms
can be derived from merging ones by swapping reads and writes. Our
goal then is to merge a set of $n$ chunks into a single reconstructed
3D image with $R$ voxels of size $b$. For simplicity, we assume that
all chunks are of identical size, that they do not overlap, and that
slices are squares and blocks are cubes.

We consider a file format where voxels are written in column-major
order, that is, for an image of dimensions (i, j, k), i is the fastest
changing dimension, followed by dimension j, followed by dimension
k. All voxels in a \emph{slice} have a the same $k$ and all voxels in
a \emph{row} have the same $j$.  NIfTI, the file format defined by the
Neuroimaging Informatics Technology Initiative
NIfTI\footnote{\url{https://nifti.nimh.nih.gov}} is such a format,
used by the vast majority of the neuroimaging community.

@@

We adopt the following notations (see Figure\ref{fig:notations}):
\begin{itemize}
\item $R=D^3$: number of voxels in the reconstructed image.
\item $b$: number of bytes per voxel (in B).
\item $n$: number of chunks (blocks or slices).
\item $m$: amount of available memory (in B).
\item $m'$: amount of used memory (in B), $m'\leq m$.
\end{itemize}
We also have the following relations:
\begin{itemize}
\item Number of slices, rows, columns in a block: $\sqrt[3]{\frac{R}{n}}=d$.
\item Number of blocks in a block row: $\sqrt[3]{n}$.
\end{itemize}

@@

A disk is characterized by its read and write rates and its seek
time. We do not distinguish between access time and seek time. We also
assume that seeks require a constant amount of time, regardless of the
position seeked to. In practice, large variations would be expected,
but modeling such variations would inevitably lead to models specific
to the hardware, file system or operating system, which we
intentionally avoid here. Likewise, in modern systems, read and write
times are greatly impacted by caching occurring at several levels,
which we do not model here. Thus, our goal is to find algorithms that
minimize the \emph{number} of seek operations.

\todo{Add ballpark values of seek time (5ms) and read/write rates.}

@@

\begin{figure}
  \centering
  \begin{minipage}[b]{0.42\columnwidth}
    \def\svgwidth{0.8\columnwidth}
    \input{figures/svg/Notations.pdf_tex}
    \caption{Notations. A \emph{block row} is shown in red. A
      \emph{block slice} is shown in blue.}
    \label{fig:notations}
  \end{minipage}
  \quad \quad \quad
  \begin{minipage}[b]{0.42\columnwidth}
    \def\svgwidth{0.8\columnwidth}
    \input{figures/svg/buffer.pdf_tex}
    \caption{Buffer used in cluster reads (d=4).  White portions in
      the buffer are not allocated. }
    \label{fig:cluster-reads-buffer}
  \end{minipage}
\end{figure}
Algorithms~\ref{algo:naive-slices} and~\ref{algo:naive-blocks} show
the naive merging methods for slices and blocks. These algorithms
actually have very different complexities even though blocks and
slices have identical sizes. Since slices are stored contiguously in
the reconstructed image, the number of seeks in
Algorithm~\ref{algo:naive-slices} is only $2n$ as $n$ seeks are
required to read the slices and $n$ seeks are required to write them:
\begin{equation}
  N_\mathrm{slices} = 2n 
\end{equation}
However,
Algorithm~\ref{algo:naive-blocks} has to do extra seeks for each row
in each slice of each block:
\begin{equation*}
N_\mathrm{blocks} = n+nd^2 
\end{equation*}
or, using $R$ and $n$ as main variables:
\begin{equation}
N_\mathrm{blocks} = n+n\left(\sqrt[3]{\frac{R}{n}}\right)^2
\end{equation}
In practice, this difference could lead to a tremendous slowdown, as
we will show later.

\begin{algorithm}
\caption{Naive merging from slices}
\label{algo:naive-slices} 
\begin{algorithmic}
  \FOR{each slice}
    \STATE read slice
    \STATE write slice in reconstructed image
  \ENDFOR      
\end{algorithmic}
\end{algorithm}
\begin{algorithm}[h]
\caption{Naive merging from blocks}
\label{algo:naive-blocks}
\begin{algorithmic}
  \FOR{each block}
    \STATE read block
    \STATE write block in reconstructed image
  \ENDFOR 
\end{algorithmic}
\end{algorithm}
%%  For slices, the total merge time is modeled as follows:
%% \begin{equation}
%% T_\mathrm{slices} = \frac{bR}{\alpha_r}+\frac{bR}{\alpha_w}+2\beta n
%% \end{equation}
%% and for blocks:
%% \begin{equation}
%%   \begin{multlined}
%%     T_\mathrm{blocks} = \frac{bR}{\alpha_r}+\frac{bR}{\alpha_w}+ 2\beta n + \\
%%     n\left({\sigma(bR/2)} + d(d-1)\sigma\left((D-d)b\right)+d\sigma\left((D^2-d^2)b\right) \right)
%%     \end{multlined}
%% \end{equation}

@@

Algorithm~\ref{algo:naive-slices} is a particular case of memory
buffering where the amount of available memory equals the maximum size
of a chunk. More buffering can be achieved when the amount of
available memory increases, as shown in
Algorithm~\ref{algo:buffered-slices}.
\begin{algorithm}[h]
  \caption{Buffered merging from slices}
  \label{algo:buffered-slices}
  \begin{algorithmic}[1]
    \STATE sorted\_slices = sort slices by increasing k values
    \STATE initialize buffer
    \FOR{i = 0 ; i \textless n ; i+=1 }
      \STATE slice = sorted\_slices[i]
      \IF{\texttt{sizeof}(buffer)+\texttt{sizeof}(slice) $\geq$ m}
        \STATE write buffer in reconstructed image
        \STATE clear buffer
      \ENDIF
      \STATE read slice and append it to buffer
    \ENDFOR
  \end{algorithmic}
\end{algorithm}
This algorithm writes in the reconstructed image using a single seek
per memory load. Therefore:
\begin{equation}
N_\mathrm{buff\_slices} =  n + \ceil*{\frac{bR}{m}}
\end{equation}
Buffered slices are straightforward to implement, however, their
extension to block merging is not easy. The remainder of this Section
presents our attempts for such a generalization.

@@

Cluster reads are the more direct extension of buffered slices to
blocks: they load multiple blocks in memory, concatenate them in a
buffer and write the buffer in the reconstructed image. Seeking is
reduced compared to naive block merging since contiguous parts of the
buffer will be written without seeking. A given
block is accessed only once.

The buffer might be a slightly complex data structure, for instance an
associative array or a Python dictionary, capable of storing multiple
disjoint sequences of contiguous bytes without having to allocate
memory for the bytes between such
sequences. Figure~\ref{fig:cluster-reads-buffer} illustrates how the
buffer would fill up for the two first blocks in a reconstructed
image, assuming that blocks are of size 4x4x4.

@@

\begin{figure}
\centering
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case1-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case2-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case3.pdf_tex}
\caption{Memory-load configurations in cluster reads, leading to
  different number of seeks. Red blocks need seeking before each of
  their rows ($d^2$ seeks). Blue blocks need seeking before each of
  their slices ($d$ seeks in total). Green blocks need only a single
  seek. Grey, dashed, transparent blocks represent the contiguous
  memory loads and are added for the sake of visualization.}
\label{fig:cluster-reads}
\end{figure}
Our cluster reads algorithm focuses on the three memory load
configurations represented in Figure~\ref{fig:cluster-reads}, that is,
the amount of memory $m'$ used by the algorithm is rounded down to the
closest number of complete blocks (case 1), of complete block rows (case
2) or of complete block slices (case 3). This is in general reasonable
since adding an incomplete row to a set of complete ones multiplies
the number of required seeks by $d$, as illustrated in
Figure~\ref{fig:avoided-configurations}-Left. In some cases though,
rounding $m$ down to $m'$ might increase the number of required
memory loads to a point that the overall number of seeks also
increases. Such cases are, however, slightly unusual and their
complete description requires extensive calculations involving modulo
arithmetic, which we felt were unwieldy to report here.
\begin{figure}
  \centering
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/incomplete-rows.pdf_tex}
\quad \quad \quad
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/overlap.pdf_tex}
\caption{Configurations that increase the number of seeks per
  memory-load and are thus deliberately avoided by cluster
  reads. Left: configuration with incomplete block rows (multiplies
  number of seeks by $d$). Right: configuration with block rows that
  overlap multiple block slices (multiplies number of seeks by $2$).}
\label{fig:avoided-configurations}
\end{figure}

Our algorithm also avoids configurations where the memory load
overlaps multiple block slices in case 2 or multiple block rows in
case 1, as such overlaps multiply the number of required seeks (see
Figure~\ref{fig:avoided-configurations}-Right).

Cluster reads are described in
Algorithm~\ref{algo:cluster-reads}. Function \texttt{switch} (line 3)
selects one of the three cases based on the amount of available memory
and the number of blocks. It returns \texttt{m'} and \texttt{case},
the identifier of the selected case. Function \texttt{check\_overlap}
(line 7) determines whether two blocks overlap multiple block slices
(case 2) or multiple block rows (case 1). For case 3 it always returns
false.  Function \texttt{sizeof} (line 8) returns the actual memory
used by its argument, including only its allocated segments in case
the argument is a buffer.
\begin{algorithm}[h]
  \caption{Buffered merging of blocks with cluster reads}
  \label{algo:cluster-reads}
  \begin{algorithmic}[1]
    \STATE sorted\_blocks = sort blocks by increasing (k,j,i)
    \STATE initialize buffer
    \STATE (m',case)=switch(m,n,R,b)
    \STATE old\_block = sorted\_blocks[0]
    \FOR{i = 0 ; i\textless n ; i+=1}
      \STATE block = sorted\_blocks[i]
      \STATE overlap = check\_overlap(block,old\_block,case)
      \IF{sizeof(buffer)+sizeof(block) $\geq$ m' \textbf{or} overlap=true}
      \STATE write buffer in reconstructed image
      \STATE clear buffer
      \STATE overlap = false
      \ENDIF
      \STATE read block and insert it in buffer
      \ENDFOR
  \end{algorithmic}
\end{algorithm}

The amount of memory used $m'$ is set as follows in each of the 3 cases:
\begin{equation*}
  m_1' = \frac{Rb}{n}\floor*{\frac{mn}{Rb}};
  m_2' = \frac{Rb}{\sqrt[3]{n}^2}\floor*{\frac{m\sqrt[3]{n}^2}{Rb}};
  m_3' = \frac{Rb}{\sqrt[3]{n}}\floor*{\frac{m\sqrt[3]{n}}{Rb}}
\end{equation*}
The number of seeks performed by cluster reads in each of the three cases is:
\begin{equation*}
  N^i_{\mathrm{CR}} = n + x_ib_i, \quad i \in \llbracket 1, 3\rrbracket,
\end{equation*}
where $x_i$ is the number of memory loads required to reconstruct the
image and $b_i$ is the number of seeks required to write the memory
load. The first $n$ seeks in the equation are required to read all the
blocks once. According to Figure~\ref{fig:cluster-reads}, we have:
\begin{equation*}
  b_1=d^2=\sqrt[3]{\frac{R}{n}}^2 \quad ; \quad b_2=d=\sqrt[3]{\frac{R}{n}} \quad ; \quad b_3=1
\end{equation*}
The
numbers of memory loads required to reconstruct the image are:
\begin{equation*}
  x_1 = \ceil*{\frac{Rb}{\sqrt[3]{n}^2m_1'}}\sqrt[3]{n}^2;
  x_2 = \ceil*{\frac{Rb}{\sqrt[3]{n}m_2'}}\sqrt[3]{n};
  x_3 = \ceil*{\frac{Rb}{m_3'}}
\end{equation*}
Because our algorithm avoids overlapping configurations, $x_1$ is
proportional to the total number of block rows in the image
($\sqrt[3]{n}^2$) and $x_2$ is proportional to the total number of
block slices ($\sqrt[3]{n}$).

@@

Multiple reads are shown in Algorithm~\ref{algo:multiple-reads}.  The
main idea of this algorithm is that blocks are read partially (line 9)
to ensure that the memory buffer only contains contiguous
bytes. Therefore, the buffer can be written continuously to the
reconstructed image, without seeking (line 13).  However, a given
block might be read multiple times, in different memory loads.

@@

\begin{algorithm}[h]
  \caption{Buffered merging of blocks with multiple reads}
  \label{algo:multiple-reads}
  \begin{algorithmic}[1]
  \STATE sorted\_blocks = sort blocks by increasing (k,j,i)
  \STATE start\_index = 0 ; end\_index=(m-1)
  \STATE write\_range = (start\_index, end\_index)
  \WHILE{end\_index \textless Rb}
    \STATE initialize buffer
    \FOR{split in sorted\_blocks}
      \IF{split has voxels in write\_range}
        \STATE split\_data = read split
        \STATE in split\_data, extract the rows in write\_range
        \STATE insert rows in buffer
      \ENDIF
    \ENDFOR
    \STATE write buffer to reconstructed\_image
    \STATE start\_index = end\_index + 1 ; end\_index += m
  \ENDWHILE

@@

As illustrated in Figure~\ref{fig:multiple-reads-cases}, we assume
that $m'$ represents an integer number $k$ of sub-rows (case 1,
$k_1<\sqrt[3]{n}$), of complete rows (case 2, $k_2<d$), of tile rows
(case 3, $k_3<\sqrt[3]{n}$), of slices (case 4, $k_4<d$) or of block
slices (case 5, $k_5<\sqrt[3]{n}$). In each of these 5 cases, we
define $v_i$ as follows:
\begin{equation*}
  v_1=db \text{ ; }  v_2=Db \text{ ; } v_3=Ddb \text{ ; } v_4=D^2b \text{ ; } v_5=D^2db
\end{equation*}
so that we have:
\begin{equation*}
k_i=\floor*{\frac{m}{v_i}} \quad \mathrm{and} \quad m_i'=k_iv_i, \quad i \in \llbracket 1,5 \rrbracket
\end{equation*}

The total number of seeks performed by multiple reads in case $i$ is:
\begin{eqnarray*}
  N^i_{\mathrm{MR}} &=& x_i + (x_i-1)b_i +b_i', \quad i \in \llbracket 1,5 \rrbracket\\
  &=& x_i \left(1+b_i\right)-b_i+b_i'
\end{eqnarray*}
where $x_i$ is the total number of memory loads, $b_i$ is the
number of blocks accessed by the first ($x_i$-1) memory loads and
$b_i'$ is the number of blocks access by the last memory load. The
first $x_i$ seeks in the equation are required to write each memory
load (1 seek per memory load). We have:
\begin{equation*}
  x_i = \ceil*{\frac{Rb}{m_i'}}, \quad i \in \llbracket 1,5 \rrbracket
\end{equation*}
and:
\begin{equation*}
b_1=k_1 ; \quad b_2=\sqrt[3]{n} ; \quad b_3 = k_3\sqrt[3]{n} ; \quad b_4=\sqrt[3]{n}^2 ; \quad b_5=k_5\sqrt[3]{n}^2
\end{equation*}
and:
\begin{eqnarray*}
b_1'=\sqrt[3]{n}D^2\text{ mod }k_1; &b_2'=b_2;   & b_3' = \sqrt[3]{n}\left( \sqrt[3]{n}D\text{ mod }k_3\right)\\
                                   &b_4'=b_4;  & b_5'=\sqrt[3]{n}^2\left( \sqrt[3]{n}\text{ mod }k_5\right) 
\end{eqnarray*}
It gives the following expression for $N_\mathrm{MR}$:
\begin{equation*}
  N_\mathrm{MR} =
\begin{cases}
  \ceil*{\frac{Rb}{m_1'}}(k_1+1)-k_1 \\
  + \left( \sqrt[3]{n}D^2 \text{ mod } k_1 \right)& \text{\hspace*{-0.9em}if } d \leq \frac{m}{b} < D\\[0.4em]

@@

\begin{itemize}
\item In MR, no seeking withing the blocks.
\item Avoid in-memory copies.
\end{itemize}

@@

For faster reading all the blocks, before the merging process start,
we read all the split’s headers to get their position information and
calculate each row’s index in the reconstructed image. We only read the
split when there is a row in that writing range (available memory),
otherwise, we skip it.

@@

% Parallel algorithms
In a real system, high-resolution images are likely to be processed on
computing clusters, for instance using software from the Hadoop
project, in particular the Hadoop Distributed File
System~\cite{shvachko2010hadoop}. In this context parallel split-and-merge algorithms
will be required, since the various blocks of a large image would be
uploaded to different disks concurrently. In the same vein,
``re-spliting'' algorithms would have to be devised in case a split
image need to be split in a different geometry. Cluster reads and
multiple reads could be used as starting points for such algorithms.

@@

% Library is available. We recommend using xxx for most cases. Same
% algo works for blocks or slices.
